I/O simulation extension \tristan{``extension'' is too general} for SimGrid framework \tristan{``SimGrid framework'' is too specific}

### 2020-06-01 paper_io_sim.tex
Some comments

\begin{document}
\title{I/O simulation extension for SimGrid framework}
\author{Hoang-Dung Do}
\maketitle

@@

	\section{Method}
		In this section, we discuss our approach to model page cache and the cache eviction strategy implemented in Linux. We also detail the design the simulators in order to generalize page cache as well as the implementation in Python and SimGrid framework. Finally, we describe some experimental scenarios to compare the results of Python simulator, baseline SimGrid simulator and the results from the real pipelines.
		\subsection{Principle of the simulator}

			When modeling the I/O time, we consider the key factors of the I/O mechanism having the biggest impacts on data read/write. The first factor taken into account is the use of the page cache in Linux kernel. In addition, the kernel controls I/O activities based on the status of the memory including the information about the cache size, the amount of available memory and dirty data. As these are managed and changed with the flushing and cache eviction mechanisms implemented in the kernel, data flushing and cache eviction must be generalized. The simulator should also make system parameters such as dirty{\_}ratio and dirty{\_}background{\_}ratio configurable since they can vary in different systems.

			The idea of our simulator is to generalize the memory, the page cache, the LRU lists, storage device and mimic the cache eviction algorithm implemented in the kernel. As for storage devices, we can leverage the storage model with disk capacity and disk bandwidth implemented in SimGrid \cite{lebre2015}. Because memory capacity and bandwidth are fixed, they can be kept as constants, while the amount of cache is dynamically allocated based on the amount of available memory. The simulator also keeps track of the LRU lists as in the kernel since they are the key component in cache eviction algorithm. However, maintaining and handling pages, which are in packet level, are non-trivial when they require significant implementation overhead and hinder the performance of the simulator. To simplify cache eviction while preserving accuracy, in our approach, we consider a sequence of continuously accessed pages of a file as a single unit, a block of data. The information of a single page in a sequence can be used as the representative for other pages in a sequence since they have the same properties including file name, last access, the page is dirty or not. Hence, instead of dealing with individual pages, we consider a sequence of continuously accessed pages as a block of data. A data block keeps the information about file name, block size, last access and a dirty flag describing whether the data is dirty or not. Since a block represents a set of pages in real systems, it can be split if some data needs to be moved between the LRU lists or evicted from cache, but blocks can not be merged as we want to preserve this level of granularity to maintain the accuracy. Having these components, we can generalize I/O activities in our simulator. When the kernel reads and writes data, it simply reads/writes data from/to disk or cache with predefined disk bandwidth or memory bandwidth. In flushing and periodical flushing, the kernel traverses through the inactive and active lists to reduce the amount of dirty data, while cache eviction algorithm deletes blocks from the inactive list. The LRU lists are updated after every I/O action.

			We create an extra layer with kernel-like functionalities between user application and SimGrid as is shown in Figure~\ref{fig:interaction}. Our new layer including two main additional classes, the \textbf{\textit{IOController}} class, which mimics the kernel I/O controlling logic, and the \textbf{\textit{MemoryManager}} class, which stores all necessary information about memory including capacity, the amount of cache, amount of memory used by applications, and manages LRU lists. Because every node has its own memory, storage devices and system configurations, there is one IOController object and one MemoryManager object created for each node. The IOController class keeps system parameters (dirty{\_}ratio, dirty{\_}background{\_}ratio), implements necessary I/O control functionalities such as read, write, flushing, and cache evicting. With these new components, instead of sending read/write requests directly to storage devices as in simulators with the current SimGrid version, user simulated applications send requests to an IOController of the node where requested data is stored. Based on the memory status in MemoryManager, the IOController orchestrates cache read/write, disk read/write, flushing and cache eviction to fulfill application requests. The time for each action is estimated and the total time of a request is returned to user applications.

@@

			\begin{algorithm}\caption{Read}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{$file\_name$}{file name}
			        	\Desc{$run\_time$}{current simulated run time}
   					\EndInput
					\State $mem\_required \gets max(0, 2* $file size$-cached\_data -$ available memory)
					\State $mem\_avail \gets max(0, 2 *$ file size$ - cached\_data -$ dirty data)
					\If {$mem\_required > mem\_avail$}
    					\State $run\_time \gets run\_time + $  flushing $(mem\_required - mem\_avail)	$
					\EndIf

					\State $free\_mem\_required \gets max(0, 2 *$ file size $- cached\_data -$ free memory)
					\If {$free\_mem\_required > 0$}
						\State evict from cache $(free\_mem\_required)$
					\EndIf

					\If {$cached\_data > 0$}
    					\State update cached blocks in LRU lists with the last access
						\If {all file data is cached}
    						\State $cache\_read\_time \gets cached\_data /$ memory read bandwidth
    						\State periodically flush $($flushing time $= cache\_read\_time)$

@@

			The algorithm to generalize file read is described in \textbf{\textit{Algorithm~\ref{alg:read}}}. In Linux systems, data can be read in two ways, read from cache with memory bandwidth and read from disk with disk bandwidth. Because total read time can be seen as a linear function of the read time in each of these two modes, we can generalize reading process as two separate and sequential phases, one is cache read, and the other is disk read. When a user application requests a file read, it calls the read method of the IOController of which a parameter is the file name. Then, IOController calculates how much memory is required to read the file based on the amount of available memory, file size, and the amount of file data cached. If there is not enough available memory, dirty data is flushed to disk to make more memory available by calling flush function with the amount of memory needed as the parameter. IOController then evicts old  blocks from the inactive list to accommodate the requested file. After that, IOController starts estimating file read time by checking if a some file data is cached. If there is cached data, the blocks containing this data are  updated in the LRU lists with the last access time. If all file data is cached, time to read from cache is added to total run time, and periodical flushing is triggered to flush dirty data during the cache read. If only a part of the file is in cache, disk read is required, this does not allow flushing and makes the cache read time neglectable. At the beginning of a disk read, dirty data is flushed \textcolor{red}{(Is this correct?)}, then data is read and added to the inactive list, disk read time is estimated with disk read bandwidth and added to total run time.

			\textbf{\textit{Algorithm~\ref{alg:write}}} describes our generalization of file write. Similar to the file read process, a file can also be written with two different bandwidths, memory bandwidth and disk bandwidth, and file write can be generalized with two modes. File can be written with memory bandwidth before the page cache is saturated, or the amount of dirty data reaches dirty\_ratio. In our simulator, dirty data can be flushed concurrently during cache write, and the maximum amount of data can be written with memory bandwidth is calculated with the formula:
			\begin{equation}
				M = \frac{(dirty\_ratio*C - d)*mw}{mw - dw}
			\end{equation}

			where:
			\begin{itemize}
				\item $M$ is the maximum amount of data can be written with memory write bandwidth
				\item $A$ is the total available memory
				\item $d$ is the amount of dirty data
				\item $mw$ is the memory write bandwidth
				\item $dw$ is the disk write bandwidth			
			\end{itemize}

@@

			Data flushing is another key function in our simulator. The input of the function is the amount of flushed data and the returned value is the simulated flushing time. To flush dirty data, IOController simply traverses the sorted inactive and active list, reduces the dirty data amount by unmarking dirty blocks until the flushed amount reached the amount passed in or there is no dirty data left in cache. Similarly, periodical flushing is the function that simulates data flushing during CPU time and cache read. However, the flushed data amount is calculated with the input which is flushing time. In both flushing functions, the last block being flushed is split if its data is not entirely flushed.

@@

			Last but not leats, we need to modified LRU lists after every I/O activity. When IOController reads the data that is not cache, the data is read and added to the inactive list. If the data is already in cache, the cached blocks are moved to the active list with the last access time updated. Finally, MemoryManager makes the LRU lists balanced by moving blocks from the longer list to the shorter and sorting the lists.

@@

			Firstly, to validate our hypothesis, we create a simple simulator independent to simulation frameworks and libraries. This enables us to evaluate the accuracy and correctness of our model in a simple use case before integrating it in SimGrid.  This simulator simulates a single thread pipeline running on a system with a single core CPU, one local storage device and all input and output files are stored locally. The results from the simulator is compared and validated with the results from a real pipeline running on a real system.

@@

			In this work, we use Python 3.7 to implement the simple simulator, SimGrid 3.25 and Wrench 1.6 for SimGrid simulators.

@@

			When designing experiments, we consider a range of scenarios that have different impacts on the results. The experiments are a number of pipelines of sequential tasks. Each task reads an input file, performs some computation, generates an output file and writes it to disk. The output file of the previous task is the input file of the next task. Thus, a pipeline has only one actual input file. The size input and output files should vary have significant impacts on system I/O behaviors. The number of pipelines should also vary in different experiments since the shared bandwidth is affected by the number of concurrent I/O processes. Different storage types, which include local and shared storage, are also taken into account to evaluate the performance of our simulator on shared file systems. Finally, we adopt a real neuroimaging pipeline to evaluate the applicability of our simulator in simulating I/O time in real applications. Because our work focuses on I/O time, we assume that CPU time is correctly modeled and use the CPU time measured in the real pipelines to setup our simulated workflows.

			We use the cluster of \textit{Big Data Infrastructure for Neuroimaging Lab} at \textit{Concordia University} to conduct our experiments. The cluster consists of 1 login node which is the control node, 8 compute nodes and 4 storage nodes connected with 2 network switches. The login node has 128 GB of RAM, 1.8 TB of storage and 385 TB of Lustre shared file system. Each compute node has 6 SSDs with the size of 450GB each, 32 cores CPU and 256 GB of RAM, while each storage node has 12 HDDs and 2 SSDs. The cluster is run on CentOS 8.1 with the Slurm Workload Manager installed. We use \textbf{\textit{atop}} and \textbf{\textit{collectl}} as tools to monitor and collect data of memory, page cache status and disk throughput on the cluster.

			\subsubsection{Expriment 1}

				In this experiment, our goal is to validate our model with a simple simulator implemented in Python. Thus, we only simulate one single thread pipeline of three tasks with the Python simulator, SimGrid simulator and run on our real system. All files are read and written locally. We use the input sizes of 20 GB, 50 GB, 75 GB and 100 GB to run the pipeline on the cluster, measure the CPU time of the pipeline with each input to simulate the pipeline with the simulators.

>>>>

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}

\begin{document}
\title{I/O simulation extension \tristan{``extension'' is too general} for SimGrid framework \tristan{``SimGrid framework'' is too specific}}
\author{Hoang-Dung Do}
\maketitle

@@

	\section{Method}
		In this section, we discuss our approach to model page cache and
		the cache eviction strategy implemented in Linux. We also detail
		the design the simulators in order to generalize page cache as well
		as the implementation in Python and SimGrid framework. Finally, we
		describe some experimental scenarios to compare the results of
		Python simulator, baseline SimGrid simulator and the results from
		the real pipelines. \tristan{It would help if you could make lines
		shorter than 80 chars, as is the common practice with code.}

		\subsection{Principle of the simulator}

			When modeling the I/O time, we consider the key factors of the
			I/O mechanism having the biggest impacts on data read/write.
			The first factor taken into account is the use of the page
			cache in Linux kernel. \tristan{Writing is unclear, there should be a sentence starting with ``The second factor is ...''} In addition, the kernel controls I/O
			activities based on the status of the memory including the
			information about the cache size, the amount of available
			memory and dirty data. As these are managed and changed with
			the flushing and cache eviction mechanisms implemented in the
			kernel, data flushing and cache eviction must be generalized \tristan{not sure what this sentence means, it should be rewritten}.
			The simulator should also make system parameters such as
			dirty{\_}ratio and dirty{\_}background{\_}ratio configurable
			since they can vary in different systems \tristan{is it important to know here?}.

			The idea of our simulator is to generalize \tristan{A simulator doesn't ``generalize'', it ``models'' or simply ``simulate''}
			 the memory, the page
			cache, the LRU lists, storage device and mimic \tristan{why a different verb for that?} the cache
			eviction algorithm implemented in the kernel. As for storage
			devices, we can leverage the storage model with disk capacity
			and disk bandwidth implemented in SimGrid~\cite{lebre2015}.
			Because memory capacity and bandwidth are fixed, they can be
			kept as constants, while the amount of cache is dynamically
			allocated based on the amount of available memory \tristan{what is the important message in this sentence?}. The
			simulator also keeps track of the LRU lists as in the kernel
			since they are the key component in cache eviction algorithm.
			\tristan{This paragraph probably needs to be broken down in several ones, 
			it seems to be mixing multiple ideas. Remember: 1 paragraph = 1 idea. The first sentence of the
			paragraph should summarize this idea.}
			However, maintaining and handling pages, which are in packet
			level, are non-trivial when they require significant
			implementation overhead and hinder the performance of the
			simulator \tristan{how do you know? if it's just a guess, you should write it as such: ``and would likely hinder ...''}.
			 To simplify cache eviction while preserving
			accuracy, in our approach, we consider a sequence of
			continuously accessed pages of a file as a single unit, a block
			of data \tristan{In general it's clearer to explain first what you did, make sure that the reader understands, and 
			contrast it next with what you didn't do and why}.
			 The information of a single page in a sequence can be
			used as the representative for other pages in a sequence since
			they have the same properties including file name, last access,
			the page is dirty or not. Hence, instead of dealing with
			individual pages, we consider a sequence of continuously
			accessed pages as a block of data. A data block keeps the
			information about file name, block size, last access and a
			dirty flag describing whether the data is dirty or not. Since a
			block represents a set of pages in real systems, it can be
			split if some data needs to be moved between the LRU lists or
			evicted from cache, but blocks can not be merged as we want to
			preserve this level of granularity to maintain the accuracy.
			Having these components, we can generalize I/O activities in
			our simulator. When the kernel reads and writes data, it simply
			reads/writes data from/to disk or cache with predefined disk
			bandwidth or memory bandwidth. In flushing and periodical
			flushing, the kernel traverses through the inactive and active
			lists to reduce the amount of dirty data, while cache eviction
			algorithm deletes blocks from the inactive list. The LRU lists
			are updated after every I/O action.

			We create an extra layer with kernel-like functionalities
			between user application and SimGrid (Figure~\ref{fig:interaction}). \tristan{This is already implementation, as it is specific to Simgrid}
			Our new layer includes two main
			classes, the \textbf{\textit{IOController}},
			which models the kernel I/O controlling logic, and the
			\textbf{\textit{MemoryManager}}, which manages LRU lists \tristan{an adjective is missing to refer unambiguously to these lists: ``page cache LRU lists?'' (here and 
			for other occurrences)} and stores all
			necessary information about memory including capacity, the
			amount of cache, amount of memory used by applications. Because every node has its own memory,
			storage devices and system configurations, there is one
			IOController object and one MemoryManager object created for
			each node. The IOController class keeps system parameters
			(dirty{\_}ratio, dirty{\_}background{\_}ratio), implements
			necessary I/O control functionalities such as read, write,
			flushing, and cache evicting. With these new components,
			instead of sending read/write requests directly to storage
			devices as in simulators with the current SimGrid version, user
			simulated applications send requests to an IOController of the
			node where requested data is stored. Based on the memory status
			in MemoryManager, the IOController orchestrates cache
			read/write, disk read/write, flushing and cache eviction to
			fulfill application requests. The time for each action is
			estimated and the total time of a request is returned to user
			applications.

@@

			\begin{algorithm}\caption{Read \tristan{Avoid using math mode for variable names}}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{$file\_name$}{file name}
						\Desc{$run\_time$}{current simulated run time}
					\tristan{Add an output}
   					\EndInput
					\State $mem\_required \gets max(0, 2* $file size$-cached\_data -$ available memory) \tristan{you should explain where available memory and cached\_data are coming from}
					\State $mem\_avail \gets max(0, 2 *$ file size$ - cached\_data -$ dirty data)
					\If {$mem\_required > mem\_avail$}
    					\State $run\_time \gets run\_time + $  flushing $(mem\_required - mem\_avail)	$
					\EndIf

					\State $free\_mem\_required \gets max(0, 2 *$ file size $- cached\_data -$ free memory)
					\If {$free\_mem\_required > 0$}
						\State evict from cache $(free\_mem\_required)$ \tristan{call a function from MemoryManager}
					\EndIf

					\If {$cached\_data > 0$}
    					\State update cached blocks in LRU lists with the last access \tristan{call a function from MemoryManager}
						\If {all file data is cached}
    						\State $cache\_read\_time \gets cached\_data /$ memory read bandwidth \tristan{memory read bandwidth should be an input}
    						\State periodically flush $($flushing time $= cache\_read\_time)$ \tristan{where does this function come from?}

@@

			\tristan{In the algo, all variables and functions should be defined:
			\begin{itemize}
				\item As inputs or class variables
				\item As class variables or functions of another class
			\end{itemize}
			Try to:
			\begin{itemize}
			\item reduce the number of lines, for instance you could always call evict\_from\_cache (please add \_s btw)
			and avoid an if statement there.
			\item streamline the logic: avoid a return statement in a second level of 
			if statements.
			\end{itemize}
			Refer to the two classes when you define and call functions. 
			}
			The algorithm to generalize file read is described in
			\textit{Algorithm~\ref{alg:read}} \tristan{(I removed the bold: avoid using bold in general.
			 You can use $\backslash$emph if you want to put emphasis on 
			something.)}. \tristan{refer to lines in Algo 1 throughout this paragraph, in brackets at the end of a sentence (line 2-4).} In Linux systems,
			data can be read in two ways, read from cache with memory
			bandwidth and read from disk with disk bandwidth. Because total
			read time can be seen as a linear function of the read time in
			each of these two modes, we can generalize reading process as
			two separate and sequential phases, one is cache read, and the
			other is disk read. When a user application requests a file
			read, it calls the read method of the IOController of which a
			parameter is the file name. Then, IOController calculates how
			much memory is required to read the file based on the amount of
			available memory, file size, and the amount of file data
			cached. If there is not enough available memory, dirty data is
			flushed to disk to make more memory available by calling flush
			function with the amount of memory needed as the parameter.
			IOController then evicts old  blocks from the inactive list to
			accommodate the requested file. After that, IOController starts
			estimating file read time by checking if a some file data is
			cached. If there is cached data, the blocks containing this
			data are  updated in the LRU lists with the last access time.
			If all file data is cached, time to read from cache is added to
			total run time, and periodical flushing is triggered to flush
			dirty data during the cache read. If only a part of the file is
			in cache, disk read is required, this does not allow flushing
			and makes the cache read time neglectable. At the beginning of
			a disk read, dirty data is flushed \textcolor{red}{(Is this
			correct?)}, then data is read and added to the inactive list,
			disk read time is estimated with disk read bandwidth and added
			to total run time.

			\textbf{\textit{Algorithm~\ref{alg:write}}} \tristan{fix consistently with Algo 1.} describes our
			generalization of file write. Similar to the file read process,
			a file can also be written with two different bandwidths,
			memory bandwidth and disk bandwidth, and file write can be
			generalized with two modes. File can be written with memory
			bandwidth before the page cache is saturated, or the amount of
			dirty data reaches dirty\_ratio. In our simulator, dirty data
			can be flushed concurrently during cache write, and the maximum
			amount of data can be written with memory bandwidth is
			calculated with the formula:
			\begin{equation}
				M = \frac{(dirty\_ratio*C - d)*mw}{mw - dw}
			\end{equation}

			where:
			\begin{itemize}
				\item $M$ is the maximum amount of data can be written with memory write bandwidth
				\item $A$ is the total available memory \tristan{Should it be C?}
				\item $d$ is the amount of dirty data
				\item $mw$ is the memory write bandwidth \tristan{Make notations consistent in the algorithms}
				\item $dw$ is the disk write bandwidth			
			\end{itemize}
			\tristan{M can be negative, add a max} \tristan{Add a sentence to explain this equation.	}
			\tristan{In math mode, variable names should be 1 char max. Avoid $dirty\_ratio$ and even $mw$. Use $m_w$ instead.}

@@

			Data flushing is another key function in our simulator. The input of the function \tristan{which function? Be explicit} is the amount of flushed data and the returned value is the simulated flushing time. To flush dirty data, IOController simply traverses the sorted inactive and active list, reduces the dirty data amount by unmarking dirty blocks until the flushed amount reached the amount passed in or there is no dirty data left in cache. Similarly, periodical flushing is the function that simulates data flushing during CPU time and cache read. However, the flushed data amount is calculated with the input which is flushing time. In both flushing functions, the last block being flushed is split if its data is not entirely flushed.

@@

			Last but not least, we need to modified LRU lists after every I/O activity. When IOController reads the data that is not cache, the data is read and added to the inactive list. If the data is already in cache, the cached blocks are moved to the active list with the last access time updated. Finally, MemoryManager makes the LRU lists balanced by moving blocks from the longer list to the shorter and sorting the lists.

@@

			Firstly, to validate our hypothesis \tristan{hypothesis has a
			strong meaning in science. ``simulation model'' would be
			better}, we create a simple simulator independent to simulation
			frameworks and libraries. This enables us to evaluate the
			accuracy and correctness of our model in a simple use case
			before integrating it in SimGrid. This simulator simulates a
			single-threaded pipeline running on a system with a single core
			CPU, one local storage device and all input and output files
			are stored locally. The results from the simulator is compared
			and validated with the results from a real pipeline running on
			a real system.

@@

			In this work, we use Python 3.7 to implement the simple
			simulator, SimGrid 3.25 and Wrench 1.6 for SimGrid simulators. \tristan{make sure to refer to the GitHub repo somewhere}

@@

			When designing experiments, we consider a range of scenarios
			that have different impacts on the results. The experiments are
			a number of pipelines of sequential tasks. Each task reads an
			input file, performs some computation, generates an output file
			and writes it to disk. The output file of the previous task is
			the input file of the next task. Thus, a pipeline has only one
			actual input file. The size input and output files should vary
			have significant impacts on system I/O behaviors. The number of
			pipelines should also vary in different experiments since the
			shared bandwidth is affected by the number of concurrent I/O
			processes. Different storage types, which include local and
			shared storage, are also taken into account to evaluate the
			performance of our simulator on shared file systems. Finally,
			we adopt a real neuroimaging pipeline to evaluate the
			applicability of our simulator in simulating I/O time in real
			applications. Because our work focuses on I/O time, we assume
			that CPU time is correctly modeled and use the CPU time
			measured in the real pipelines to setup our simulated
			workflows. \tristan{This paragraph mixes too many things: you
			should make one on pipelines, one on multi-threaded vs
			single-treaded, one on data (file creation and size), a separate one on infrastructure (storage),
			etc}

			We use the cluster of \textit{Big Data Infrastructure for
			Neuroimaging Lab} at \textit{Concordia University} to conduct
			our experiments. The cluster consists of 1 login node which is
			the control node, 8 compute nodes and 4 storage nodes connected
			with 2 network switches. The login node has 128 GB of RAM, 1.8
			TB of storage and 385 TB of Lustre shared file system. Each
			compute node has 6 SSDs with the size of 450GB each, 32 cores
			CPU and 256 GB of RAM, while each storage node has 12 HDDs and
			2 SSDs. The cluster is run on CentOS 8.1 with the Slurm
			Workload Manager installed. We use \textbf{\textit{atop}} and
			\textbf{\textit{collectl}} as tools to monitor and collect data
			of memory, page cache status and disk throughput on the
			cluster. \tristan{You should report file system types and versions, CPU and disk types, network type and bandwidth, Lustre configuration.}

			\subsubsection{Experiment 1 \tristan{Give experiments a more meaningful name than a number. For instance: ``Single-threaded evaluation''}}

				In this experiment, our goal is to validate our model with
				a simple simulator implemented in Python \tristan{If that was your only goal you wouldn't need simgrid at all in this experiment}. Thus, we only
				simulate one single thread pipeline of three tasks with the
				Python simulator, SimGrid simulator and run on our real
				system. All files are read and written locally. We use the
				input sizes of 20 GB, 50 GB, 75 GB and 100 GB to run the
				pipeline on the cluster, measure the CPU time of the
				pipeline with each input to simulate the pipeline with the
				simulators.
